The Name of the Title Is Hope

### 2024-12-26 paper.tex
typography, esp. using booktabs for tables

 Most relevant to this work is the scheduling behavior of the GPU. Scheduling is divided into a workgroup-level scheduler\footnote{For example, NVIDIA's \emph{GigaThread Engine} or AMD's \emph{Asynchronous Compute Engines}}, which manages kernel launches and maps workgroups to multiprocessors, and the aforementioned subgroup scheduler, responsible for selecting which of the currently resident subgroups for execution. Although contemporary GPU hardware supports workgroup preemption to prioritize real-time tasks or manage multi-process workloads\cite{}(NVIDIA Volta,AMD RDNA), this capability does not extend to the GPU programming model. Once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context switch. At the subgroup scheduling level, \emph{fairness}—how evenly execution resources are distributed among threads—and \emph{progress guarantees}—ensuring that all threads or subgroups eventually make forward progress—are the most salient issues. As we will discuss in \emph{Why does \emph{Chained-Scan} Rely on Forward Progress Guarantees} (latex section link here), these factors have profound implications for kernel performance and correctness.

\subsection{The Scan Primitive}
The study of \emph{parallel prefix computation networks} traces back to the design of carry-lookahead adder circuits and earlier work\cite{10.1145/322217.322232, 5219801}. The term \emph{scan}, introduced by Iverson in the APL programming language, is often used interchangeably with \emph{prefix computation}\cite{IVERSON}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid\cite{}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence. If the subset includes the \( n \)-th element, it is called \textit{inclusive}; if it excludes the \( n \)-th element, it is called \textit{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
x = [x_1, x_2, x_3, \dots, x_n]
\]
\[
\text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]

@@

 Scans are characterized by their \emph{depth} $d$, the number of parallel work steps required, and \emph{size} $s$, the total number of $\oplus$ operations performed. Snir proved that for a network size $n$, depth and size are related by $s + d \ge 2n - 2$ \cite{}. This equation highlights a fundamental tradeoff: as depth decreases to improve parallelism, the size (i.e., total work) increases proportionally, and vice versa. Likewise, Harris introduces a circuit-based, three-dimensional taxonomy of scans, relating \emph{fanout} $f$, \emph{wire-tracks} $t$, and \emph{depth} by $f + t + d = \log_2n - 1$ \cite{}. Here, fanout denotes the number of simultaneous outputs from a single node, while wire-tracks represent the number of parallel connections, broadly analogous to overall network size. The classic scans map to maximal points in this three-dimensional space: maximal tracks Kogge-Stone/Hillis-Steele\cite{}, maximal fanout Sklansky\cite{}, and maximal depth Brent-Kung\cite{}. Furthermore, the associative property of $\oplus$ enables the hybridization of different scan network structures. In Harris’s taxonomy, these hybrid scans correspond to intermediate points between the maximal designs: depth-fanout Ladner-Fischer\cite{}, depth-tracks Hans-Carlson\cite{}, and fanout-tracks Knowles\cite{}.

 \begin{table}[h]
  \centering
  \caption{Comparison of Different Scan Networks}
  \label{tab:scan_comparison_exact}
  \begin{tabular}{|l|c|c|}
  \hline
  \textbf{Network}       & \textbf{Size}        & \textbf{Miminum Depth Variant} \\ \hline
  \textbf{Serial}           & $n$           & $n - 1$            \\ \hline
  \textbf{Kogge--Stone}     & $O(n \log n)$ & $\log_2(n)$        \\ \hline
  \textbf{Sklansky}         & $O(n \log n)$ & $\log_2(n)$        \\ \hline
  \textbf{Brent--Kung}      & $O(n)$        & $2 \log_2(n) - 1$  \\ \hline
  \textbf{Han--Carlson}     & $O(n)$        & $\log_2(n) + \Big\lfloor \tfrac{\log_2(n)}{2} \Big\rfloor - 1$ \\ \hline
  \textbf{Knowles}          & $O(n)$        & $\log_2(n)$        \\ \hline
  \textbf{Ladner--Fischer}  & $O(n)$        & $\log_2(n)$        \\ \hline
  \end{tabular}
\end{table}
 \footnote{Hybrid scans offer varying degrees of hybridization; for simplicity, we report only the size of the minimal-depth variant.}
 For our purposes, we are primarily concerned about depth and size. Notably, Fich proved that with unbounded fanout, the Ladner-Fischer network has both minimal depth, $\lceil \log_2n \rceil$ and optimal size, $O(n)$ \cite{}. While circuit engineering tends to favor minimal depth networks---Kogge-Stone, Hans-Carlson---for minimal latency, the task sizes of scan algorithms frequently exceed the number of available physical processors. This makes minimal-size networks, like Brent–Kung and Sklansky, more practical for algorithmic implementation. Within computer science, Blelloch formalized scan in the PRAM model and popularized their use as a parallel compute primitive \cite{}. Horn was the first to adapt scan for GPU use and Sengupta et al. were the first to adapt scan to CUDA \cite{}.

@@

\subsection{Missing Synchronization Primitive?}
 In a scan, the reduction at each element is dependent on the reduction of preceding elements. Thus, a serial dependency is created between workgroups whenever the number of elements in a scan operation exceeds the maximum size that can be processed by a single workgroup. Historically, hardware vendors have only recently begun to formally support inter-workgroup synchronization. Prior to the introduction of \emph{thread block clusters} on NVIDIA's Hopper architecture\cite{}, we are unaware of any GPU programming framework\footnote{We are not aware of any prior inter-workgroup level barrier primitive in CUDA, OpenCL, GLSL, HLSL, or Metal.} that provided inter-workgroup synchronization primitives. Instead, the earlier \emph{Scan-then-Propagate} and \emph{Reduce-then-Scan} architectures rely on kernel launches to act as inter-workgroup synchronization points\cite{}.

@@

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
 Although NVIDIA formalized FPG down to the subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs\cite{10.1145/3485508}. Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time. It's also unclear why the test did not trigger Metal's timeout device recovery}, and at worst, risks the afformentioned TDR and subsequent program crash. This risk is further exacerbated by the fact that, no major graphics API currently provides developers an FPG hardware capability query\footnote{To the best of our knowledge D3D12, Vulkan, and Metal do not have a query for FPG}. As a result, when implementing scan operations in a shading environment—where cross-vendor, cross-architecture portability is essential—uncertainty surrounding FPG capability and its catastrophic failure mode compels developers to fall back to the older, slower \emph{Reduce-then-Scan} approach.

@@

\subsection{Intra-Workgroup Implementation}
 (TODO: Try Reduce-Then-Scan on low spec hardware.)
 Our intra-workgroup implementation is a variation of the \emph{3D-Matrix Scan} (3DMS) introduced by Yan et al.\cite{}, with adaptations to enhance portability. Similar to 3DMS, we split the work tile into non-overlapping blocks, which are distributed to each subgroup. However, unlike CUDA, WGSL lacks a \emph{volatile} qualifier for shared memory, which prevents us from loading items into shared memory and transposing them without inserting a barrier. However, placing a barrier disrupts latency hiding by separating the high-latency global memory load from the subsequent transpose and scan computation. To address this, we load data directly into registers and \emph{vectorize} the input. This approach preserves a \emph{coalesced} but vector-strided loading pattern. For each vector, a thread performs a serial scan across the vector, then participates in a subgroup scan. Although this method results in less efficient subgroup scan computation, $\frac{p}{vs}$ subgroup scans versus one, it elides the barrier and significantly reduces the kernel's shared memory footprint. Transposing in shared memory would require a work-tile-sized shared memory allocation, whereas our approach relies more heavily on registers, which are generally more abundant than shared memory, especially on low-spec hardware. As a result, we consider this tradeoff worthwhile.

 Once completed, each subgroup posts its reduction into shared memory, then participates in a workgroup-wide scan. Because the WebGPU specification supports subgroup sizes $s \text{ where } s = 2^k, \, k \in [2, 7]$, and on some hardware, the subgroup size can vary between kernel launches, it is imperative to use a scan implementation which can accomodate all subgroup sizes. (More) We use a log base $s$ Ladner-Fischer scan with Merrill-Grimshaw conflict avoidance that embeds Kogge-Stone subgroup scans\cite{}. This network has several advantages: minimal depth $log_s n$; asymptotically optimal $O(n)$ size; $s$-way conflict (seems bad, but actually turns into broadcast?).

@@

 \subsubsection{Size (Work) Efficiency}
 Recall that for a subgroup of size $s$, the Kogge-Stone scan has a size complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
 \begin{equation}
  \begin{aligned}[b]
      \text{Work} &= \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}} 
      + 
      \underbrace{\sum_{k=1}^{\lceil \log_s n \rceil}}_{\text{iterations}}
      \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}\frac{n}{s^k}}_{\text{calls per iteration}} 
      \cdot 
      \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
      &= n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k} \\
      &= O\left(n + n \log_2 s\right) \\
      &= O(n \log_2 s).
  \end{aligned}
 \end{equation}

 \subsubsection{Depth}
 Recall that a Kogge-Stone scan has a depth of $\log_2 s$ Thus, the total depth of the scan is:
 \begin{equation}
  \begin{aligned}[b]
      \text{Depth} &= \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}} 
      \cdot \underbrace{\log_2 s}_{\text{depth per iteration}} \\
      &= \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
      &= \log_2 n + O(1).
  \end{aligned}
 \end{equation}
 Minimal depth is preserved regardless of the subgroup size. However, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration. Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.
\subsection{Full Algorithm}

@@

 Fairness has been a priority in NVIDIA hardware since at least the Tesla architecture\cite{}, but the extent to which it is implemented and prioritized by other vendors remains unclear. While our results on the least fair hardware demonstrate that \emph{Decoupled-Fallback} remains performant, this performance may not extend to scans involving more complex binary reduction operators or less capable hardware.
\begin{acks}
\end{acks}

>>>>

Most relevant to this work is the scheduling behavior of the GPU. Scheduling is divided into a workgroup-level scheduler\footnote{For example, NVIDIA's \emph{GigaThread Engine} or AMD's \emph{Asynchronous Compute Engines}}, which manages kernel launches and maps workgroups to multiprocessors, and the aforementioned subgroup scheduler, responsible for selecting which of the currently resident subgroups for execution. Although contemporary GPU hardware supports workgroup preemption to prioritize real-time tasks or manage multi-process workloads~\cite{}(NVIDIA Volta,AMD RDNA), this capability does not extend to the GPU programming model. Once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context switch. At the subgroup scheduling level, \emph{fairness}—how evenly execution resources are distributed among threads—and \emph{progress guarantees}—ensuring that all threads or subgroups eventually make forward progress—are the most salient issues. As we will discuss in \emph{Why does \emph{Chained-Scan} Rely on Forward Progress Guarantees} (latex section link here), these factors have profound implications for kernel performance and correctness.

\subsection{The Scan Primitive}
The study of \emph{parallel prefix computation networks} traces back to the design of carry-lookahead adder circuits and earlier work~\cite{10.1145/322217.322232, 5219801}. The term \emph{scan}, introduced by Iverson in the APL programming language, is often used interchangeably with \emph{prefix computation}~\cite{IVERSON}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the stack monoid~\cite{}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence. If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]

@@

Scans are characterized by their \emph{depth} $d$, the number of parallel work steps required, and \emph{size} $s$, the total number of $\oplus$ operations performed. Snir proved that for a network size $n$, depth and size are related by $s + d \ge 2n - 2$~\cite{}. This equation highlights a fundamental tradeoff: as depth decreases to improve parallelism, the size (i.e., total work) increases proportionally, and vice versa. Likewise, Harris introduces a circuit-based, three-dimensional taxonomy of scans, relating \emph{fanout} $f$, \emph{wire-tracks} $t$, and \emph{depth} by $f + t + d = \log_2n - 1$~\cite{}. Here, fanout denotes the number of simultaneous outputs from a single node, while wire-tracks represent the number of parallel connections, broadly analogous to overall network size. The classic scans map to maximal points in this three-dimensional space: maximal tracks Kogge-Stone/Hillis-Steele~\cite{}, maximal fanout Sklansky~\cite{}, and maximal depth Brent-Kung~\cite{}. Furthermore, the associative property of $\oplus$ enables the hybridization of different scan network structures. In Harris’s taxonomy, these hybrid scans correspond to intermediate points between the maximal designs: depth-fanout Ladner-Fischer~\cite{}, depth-tracks Hans-Carlson~\cite{}, and fanout-tracks Knowles~\cite{}.

\begin{table}[h]
  \centering
  \caption{Comparison of Different Scan Networks}
  \label{tab:scan_comparison_exact}
  \begin{tabular}{lcc}
    \toprule
    \textbf{Network}         & \textbf{Size} & \textbf{Miminum Depth Variant}                                 \\ \midrule
    \textbf{Serial}          & $n$           & $n - 1$                                                        \\
    \textbf{Kogge--Stone}    & $O(n \log n)$ & $\log_2(n)$                                                    \\
    \textbf{Sklansky}        & $O(n \log n)$ & $\log_2(n)$                                                    \\
    \textbf{Brent--Kung}     & $O(n)$        & $2 \log_2(n) - 1$                                              \\
    \textbf{Han--Carlson}    & $O(n)$        & $\log_2(n) + \Big\lfloor \tfrac{\log_2(n)}{2} \Big\rfloor - 1$ \\
    \textbf{Knowles}         & $O(n)$        & $\log_2(n)$                                                    \\
    \textbf{Ladner--Fischer} & $O(n)$        & $\log_2(n)$                                                    \\ \bottomrule
  \end{tabular}
\end{table}
\footnote{Hybrid scans offer varying degrees of hybridization; for simplicity, we report only the size of the minimal-depth variant.}
For our purposes, we are primarily concerned about depth and size. Notably, Fich proved that with unbounded fanout, the Ladner-Fischer network has both minimal depth, $\lceil \log_2n \rceil$ and optimal size, $O(n)$~\cite{}. While circuit engineering tends to favor minimal depth networks---Kogge-Stone, Hans-Carlson---for minimal latency, the task sizes of scan algorithms frequently exceed the number of available physical processors. This makes minimal-size networks, like Brent–Kung and Sklansky, more practical for algorithmic implementation. Within computer science, Blelloch formalized scan in the PRAM model and popularized their use as a parallel compute primitive~\cite{}. Horn was the first to adapt scan for GPU use and Sengupta et al.\ were the first to adapt scan to CUDA~\cite{}.

@@

\subsection{Missing Synchronization Primitive?}
In a scan, the reduction at each element is dependent on the reduction of preceding elements. Thus, a serial dependency is created between workgroups whenever the number of elements in a scan operation exceeds the maximum size that can be processed by a single workgroup. Historically, hardware vendors have only recently begun to formally support inter-workgroup synchronization. Prior to the introduction of \emph{thread block clusters} on NVIDIA's Hopper architecture~\cite{}, we are unaware of any GPU programming framework\footnote{We are not aware of any prior inter-workgroup level barrier primitive in CUDA, OpenCL, GLSL, HLSL, or Metal.} that provided inter-workgroup synchronization primitives. Instead, the earlier \emph{Scan-then-Propagate} and \emph{Reduce-then-Scan} architectures rely on kernel launches to act as inter-workgroup synchronization points~\cite{}.

@@

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
Although NVIDIA formalized FPG down to the subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs~\cite{10.1145/3485508}. Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time. It's also unclear why the test did not trigger Metal's timeout device recovery}, and at worst, risks the afformentioned TDR and subsequent program crash. This risk is further exacerbated by the fact that, no major graphics API currently provides developers an FPG hardware capability query\footnote{To the best of our knowledge D3D12, Vulkan, and Metal do not have a query for FPG}. As a result, when implementing scan operations in a shading environment—where cross-vendor, cross-architecture portability is essential—uncertainty surrounding FPG capability and its catastrophic failure mode compels developers to fall back to the older, slower \emph{Reduce-then-Scan} approach.

@@

\subsection{Intra-Workgroup Implementation}
(TODO: Try Reduce-Then-Scan on low spec hardware.)
Our intra-workgroup implementation is a variation of the \emph{3D-Matrix Scan} (3DMS) introduced by Yan et al.~\cite{}, with adaptations to enhance portability. Similar to 3DMS, we split the work tile into non-overlapping blocks, which are distributed to each subgroup. However, unlike CUDA, WGSL lacks a \emph{volatile} qualifier for shared memory, which prevents us from loading items into shared memory and transposing them without inserting a barrier. However, placing a barrier disrupts latency hiding by separating the high-latency global memory load from the subsequent transpose and scan computation. To address this, we load data directly into registers and \emph{vectorize} the input. This approach preserves a \emph{coalesced} but vector-strided loading pattern. For each vector, a thread performs a serial scan across the vector, then participates in a subgroup scan. Although this method results in less efficient subgroup scan computation, $\frac{p}{vs}$ subgroup scans versus one, it elides the barrier and significantly reduces the kernel's shared memory footprint. Transposing in shared memory would require a work-tile-sized shared memory allocation, whereas our approach relies more heavily on registers, which are generally more abundant than shared memory, especially on low-spec hardware. As a result, we consider this tradeoff worthwhile.

Once completed, each subgroup posts its reduction into shared memory, then participates in a workgroup-wide scan. Because the WebGPU specification supports subgroup sizes $s \text{ where } s = 2^k, \, k \in [2, 7]$, and on some hardware, the subgroup size can vary between kernel launches, it is imperative to use a scan implementation which can accomodate all subgroup sizes. (More) We use a log base $s$ Ladner-Fischer scan with Merrill-Grimshaw conflict avoidance that embeds Kogge-Stone subgroup scans~\cite{}. This network has several advantages: minimal depth $log_s n$; asymptotically optimal $O(n)$ size; $s$-way conflict (seems bad, but actually turns into broadcast?).

@@

\subsubsection{Size (Work) Efficiency}
Recall that for a subgroup of size $s$, the Kogge-Stone scan has a size complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
\begin{equation}
  \begin{aligned}[b]
    \text{Work} & = \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}}
    +
    \underbrace{\sum_{k=1}^{\lceil \log_s n \rceil}}_{\text{iterations}}
    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}\frac{n}{s^k}}_{\text{calls per iteration}}
    \cdot
    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
                & = n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k}       \\
                & = O\left(n + n \log_2 s\right)                                                 \\
                & = O(n \log_2 s).
  \end{aligned}
\end{equation}

\subsubsection{Depth}
Recall that a Kogge-Stone scan has a depth of $\log_2 s$ Thus, the total depth of the scan is:
\begin{equation}
  \begin{aligned}[b]
    \text{Depth} & = \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}}
    \cdot \underbrace{\log_2 s}_{\text{depth per iteration}}                           \\
                 & = \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
                 & = \log_2 n + O(1).
  \end{aligned}
\end{equation}
Minimal depth is preserved regardless of the subgroup size. However, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration. Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.
\subsection{Full Algorithm}

@@

Fairness has been a priority in NVIDIA hardware since at least the Tesla architecture~\cite{}, but the extent to which it is implemented and prioritized by other vendors remains unclear. While our results on the least fair hardware demonstrate that \emph{Decoupled-Fallback} remains performant, this performance may not extend to scans involving more complex binary reduction operators or less capable hardware.
\begin{acks}
\end{acks}
